---
title: "COVID-19 mortality in England and Wales"
subtitle: "Based on the ONS data"
author: Kasia Kedzierska
output:
  html_document:
    df_print: paged
    code_folding: hide
    theme: paper
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---

## Disclaimer
As probably many of the data scientist, I wanted to ease my anxiety caused by the COVID-19. I decided to look at the data related to the pandemic.  I decided not to look at the daily cases as [John Burn-Murdoch](https://twitter.com/jburnmurdoch) from [Financial Times](https://www.ft.com/coronavirus-latest) and the team [Our World in Data](https://ourworldindata.org/the-covid-19-pandemic-slide-deck) are doing a great job. The reasons why I decided to look at deaths are explained in the Motivation section. 

I am a **computational biologist** and all the plots I am making here are just *observations*. I am not claiming any conclusions, nor do I do any predictions. All the code is open source, and free for anyone to use, available on [github](https://github.com/kzkedzierska/COVID-19_England_Wales).

Also, very warm thanks to [Kaspar](https://kasparmartens.rbind.io/) for his help with decisions regarding color schemes and his helpful advice.

## Motivation

I decided to look into the mortality rates because I wanted to put the numbers reported by the government in some context. My first question was whether the deaths reported as COVID-19 deaths were a subset of deaths one could expect to see this year? How much COVID-19 increases the death toll? And finally, how does the age distribution of fatalities look?

The data I plotted below shades a bit of light on those topics, but for a more in-depth answer, I am afraid we will have to wait for long. Nevertheless, I still think the following is quite interesting. If not for others, then for the sole reason of looking at the data in almost real-time.

## Setup 

Loading up the necessary packages (installing them if they are missing). Setting up defaults and tweaking the plotting defaults.

```{r, warning=FALSE, error=FALSE, message=FALSE}
# Load packages
cran_pkgs <- c("tidyverse", "ggsci", "patchwork", 
               "readxl", "knitr", "ggforce", "lubridate")
source("./scripts/load_pkgs.R")
load_packages(cran_pkgs = cran_pkgs)

# Set defaults for plotting
theme_set(theme_bw() + theme(legend.position = "bottom"))

# Don't print warnings
knitr::opts_chunk$set(warning=FALSE, message=FALSE)

# Update date of the data
data_up_to <- ymd("2020-04-24")
```
## Data 

Data up to: `r ymd(data_up_to)`

Data is coming from [Office for National Statistics, *Deaths registered weekly in England and Wales, provisional* dataset](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/weeklyprovisionalfiguresondeathsregisteredinenglandandwales). For now I am downloading the data manually into the `data` directory. It would be great to download it directly.

### Weekly figures 

Both 2020 and 2019 figures for comparison of the trend. The 2020 will be changed each week, while 2019 remains stable. 

*Note:* Unfortunately due to formatting of the data the following code is not entirely reproducible as there are parts of it that would have to be adjusted with each newiteration of the spreadsheet. 

```{r, warning=FALSE, message=FALSE}

df_2019 <- 
  read_xls("./data/publishedweek522019.xls", 
           sheet = "Weekly figures 2019", skip = 3)[c(4,5,9),-c(1,2)] %>%
  mutate(value = c("total_deaths", "total_deaths_ave5", "respiratory_deaths")) %>%
  mutate_at(vars(-value), as.numeric) %>% 
  pivot_longer(names_to = "week", 
               values_to = "number_of_deaths", -value) %>%
  mutate(year = "2019")

df_2015 <-
  read_xls("./data/publishedweek2015.xls", 
           sheet = "Weekly Figures 2015", skip = 3)[c(4,5,9),-c(1)] %>%
  mutate(value = c("total_deaths", "total_deaths_ave5", "respiratory_deaths")) %>%
  mutate_at(vars(-value), as.integer) %>% 
  pivot_longer(names_to = "week", 
               values_to = "number_of_deaths", -value) %>%
  mutate(year = "2015", 
         week = as.integer(week),
         week_ = ymd("2020-01-03") + 7 * (week-1)) 

current_2020_spreadsheet <- "./data/publishedweek172020.xlsx"
sheets_2020 <- readxl::excel_sheets(current_2020_spreadsheet)

df_2020_ <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[5], skip = 4)[c(4,6,13,14),-c(1,2)] %>%
  mutate(value2 = c("total_deaths", "total_deaths_ave5",
                   "respiratory_deaths", "covid_deaths"),
         value = c("total_deaths", "total_deaths_ave5",
                   "respiratory_deaths", "respiratory_deaths")) %>%
  mutate_at(vars(-value, -value2), as.numeric) %>% 
  pivot_longer(names_to = "week", 
               values_to = "number_of_deaths", 
               -c(value, value2))

df_2020 <- 
  df_2020_ %>% 
  select(-value2) %>%
  mutate(year = "2020") %>%
  filter(!is.na(number_of_deaths))

df_all <- bind_rows(list(df_2019, df_2020)) %>%
  mutate(week = as.numeric(week), 
         week_ = ymd("2020-01-03") + 7 * (week-1)) %>%
  group_by(value, week, week_, year) %>%
  summarise(number_of_deaths = sum(number_of_deaths)) %>%
  ungroup()
```

### COVID-19 specific datasets

```{r}
place_of_death <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[grepl("Place", sheets_2020)], 
            skip = 4)[14:20,1:7] %>%
  rename(place = 1) %>%
  pivot_longer(names_to = "country", 
               values_to = "number_of_deaths",
               c(-place)) %>%
  mutate(country = rep(c(rep("England and Wales", 2), rep("England", 2), rep("Wales", 2)), 7),
         cause = rep(rep(c("Total deaths", "COVID-19 deaths"), 3), 7)) %>%
  mutate(number_of_deaths = as.numeric(number_of_deaths))

actual_vs_reported <-
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[str_detect(sheets_2020, regex("E&W comparisons", ignore_case = TRUE))], skip = 4, 
            col_types = c("date", rep("numeric", 8)))[, 1:9] %>%
  filter(!is.na(Date)) %>%
  pivot_longer(names_to = "source",
               values_to = "number_of_deaths",
               -Date)

place_of_death_long <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[grepl("Place", sheets_2020)], 
            skip = 4)[4:9,] %>%
  rename(place = 1) %>%
  pivot_longer(names_to = "cause",
               values_to = "number_of_deaths",
               -place) %>%
  mutate(cause = rep(c("Total deaths", "COVID-19 deaths"), n()/2),
         country = rep(c(rep("England and Wales", 2), 
                         rep("England", 2), 
                         rep("Wales", 2)), n()/6),
         week = rep(rep(11:(10 + n() / 36), each = 6), 6))

age_distribution <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[str_detect(sheets_2020, 
                                           regex("covid-19 .* occurrences", 
                                                 ignore_case = TRUE))], 
            skip = 4)[c(7:26,29:48,51:70), -1] %>%
  mutate(sex = c(rep("Both", 20), rep("Males", 20), rep("Females", 20))) %>%
  rename(age_group = `...2`) %>%
  mutate_at(vars(-age_group, -sex), as.numeric) %>%
  pivot_longer(names_to = "week",
               values_to = "number_of_deaths",
               cols = -c(age_group, sex)) %>% 
  replace_na(list(number_of_deaths = 0)) %>%
  group_by(sex, week) %>%
  filter(sum(number_of_deaths, na.rm = TRUE) > 0)

age_levels <- age_distribution$age_group %>% unique()
age_distribution <-
  age_distribution %>%
  ungroup() %>%
  mutate(week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  mutate(age_group = factor(age_group, levels = age_levels))

age_distribution_all <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[str_detect(sheets_2020, 
                                           regex("Weekly figures 2020", 
                                                 ignore_case = TRUE))], 
            skip = 4)[c(17:36,39:58,61:80), -1] %>%
  mutate(sex = c(rep("Both", 20), rep("Males", 20), rep("Females", 20))) %>%
  rename(age_group = 1) %>%
  mutate_at(vars(-age_group, -sex), as.numeric) %>%
  pivot_longer(names_to = "week",
               values_to = "number_of_deaths",
               cols = -c(age_group, sex)) %>% 
  replace_na(list(number_of_deaths = 0)) %>%
  group_by(sex, week) %>%
  filter(sum(number_of_deaths, na.rm = TRUE) > 0) %>%
  ungroup() %>%
  mutate(week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  mutate(age_group = factor(age_group, levels = age_levels))
```


## Mortality in the context of past years

### All weekly deaths in England and Wales

Still very early, but looks like the overall weekly numbers are higher than last year, and the past 5 years.

```{r, fig.height=8, fig.width=7}
total <- 
  df_all %>%
  filter(value == "total_deaths") %>%
  ggplot(aes(x = week_, y = number_of_deaths, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_smooth(data = filter(df_all, value == "total_deaths_ave5" & year == "2019"),
              color = "grey60", se = FALSE) +
  facet_zoom(xy = week_ < data_up_to + 1, 
             horizontal = FALSE, show.area = FALSE) +
  labs(y = "Number of deaths", 
       title = "Weekly total deaths  [grey = 5 year average]") + 
  theme(axis.title.x = element_blank()) +
  scale_fill_aaas()
total
```

### Respiratory deaths

COVID-19 deaths are increasing the number of repiratory deaths. 

```{r, fig.height=8, fig.width=7}
upper_limit <-
  df_all %>%
  filter(value == "respiratory_deaths") %>%
  pull(number_of_deaths) %>%
  max() + 100
respiratory <-
  df_all %>%
  filter(value == "respiratory_deaths") %>%
  ggplot(aes(x = week_, y = number_of_deaths, 
             group = year, color = year, shape = year)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_zoom(xy = week_ < data_up_to + 1, 
             horizontal = FALSE, show.area = FALSE) +
  scale_color_aaas() +
  labs(y = "Number of deaths", title = "Weekly respiratory deaths") + 
  theme(axis.title.x = element_blank()) +
  coord_cartesian(ylim = c(0, upper_limit))
respiratory
```

However, the difference is not yet visible on the cummulative scale. 

```{r, fig.height=8, fig.width=7}
cumulative_respiratory <- 
  df_all %>%
  arrange(week) %>%
  group_by(value, year) %>%
  mutate(cum = cumsum(number_of_deaths)) %>%
  filter(value == "respiratory_deaths") %>%
  ungroup() %>% 
  ggplot(aes(x = week_, y = cum, color = year, shape = year)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_zoom(xy = week_ < data_up_to + 1, 
             horizontal = FALSE, show.area = FALSE) +
  scale_color_aaas() +
  labs(y = "Number of deaths", shape = "Year",
       color = "Year", title = "Cumulative respiratory deaths") + 
  theme(axis.title.x = element_blank()) 
cumulative_respiratory
```

COVID-19 deaths increase exponentailly, while the other respiratory deths seem to be stable.

As per [BBC plot](https://www.bbc.co.uk/news/health-52361519) I added the 2015 respiratory deaths, when the worst flu outbreak took place.

```{r}
df_2020_ %>%
  mutate(year = "2020",
         value2 = gsub("respiratory_deaths", "other respiratory deaths", value2),
         value2 = str_replace(value2, "covid_deaths", "COVID-19 deaths"),
         week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  bind_rows((df_2015 %>%
               mutate(value2 = ifelse(value == "respiratory_deaths", 
                                      "respiratory deaths in 2015", value)))) %>%
  filter(value == "respiratory_deaths") %>%
  filter(!is.na(number_of_deaths), 
         week_ < data_up_to + 1) %>%
  ggplot(aes(x = week_, y = number_of_deaths, color = value2, shape = value2)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_color_uchicago() +
  labs(y = "Number of deaths", title = "Weekly respiratory deaths",
       color = "Cause of death", shape = "Cause of death") + 
  theme(axis.title.x = element_blank(), legend.position = "bottom")
```

```{r, fig.width=7, fig.height=12, eval=FALSE}
total / respiratory / cumulative_respiratory + plot_layout(guides = "collect")
ggsave("./plots/up_to200327.png")
```

## COVID-19 specific stats

### Number of all COVID-19 related deaths

In this plot, I am looking at the difference in total number of deaths reported in the period since the first COVID-19 related death was reported and the 5-year average. 

We see first evidence of UK being past the peak, or in the plateau phase. 

```{r}
col_scheme <- pal_uchicago()(5)[c(1:2,5)]
names(col_scheme) <- c("increasing phase", 
                       "decreasing phase", 
                       "deaths below average")

df_2020_ %>% 
  filter(!is.na(number_of_deaths)) %>% 
  group_by(week) %>% 
  filter(any((value2 == "covid_deaths") & (number_of_deaths > 0))) %>%
  ungroup() %>%
  filter(value2 %in% c("total_deaths", "total_deaths_ave5")) %>%
  select(-value) %>%
  pivot_wider(names_from = value2, 
              values_from = number_of_deaths) %>%
  mutate(difference = total_deaths - total_deaths_ave5,
         fract = difference / total_deaths_ave5,
         week_ = lubridate::ymd("2020-01-03") + 7 * (as.numeric(week) - 1),
         group = ifelse(difference < 0,
                        "deaths below average",
                        ifelse(difference > lag(difference),
                               "increasing phase",
                               "decreasing phase"))) %>%
  ggplot(aes(week_, difference, fill = group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = col_scheme) +
  theme(legend.title = element_blank(), axis.title.x = element_blank()) +
  labs(y = "Number of deaths",
       title = "COVID-19 excess deaths",
       subtitle = "Difference in deaths over the 5-year average since first COVID-19 death reported")

```


### Place of deaths
```{r}
fract_place_of_death <- 
  place_of_death %>% 
  filter(place != "All deaths",
         country == "England and Wales") %>%
  mutate(place = gsub("Hospital \\(acute or community, not psychiatric\\)", 
                      "Hospital", place),
         place = gsub("Other communal establishment", 
                      "Other communal\nestablishment", place)) %>%
  group_by(cause) %>%
  mutate(fract = round(number_of_deaths / sum(number_of_deaths) * 100, 2))
```


As of `r format(data_up_to, '%b, %d')`, majority of the deaths are reported in hospitals. However `r 100 - round((fract_place_of_death %>% filter(place == "Hospital", cause == "COVID-19 deaths") %>% pull(fract)), 0)`% of deaths occur outside of hospitals.

```{r}
fract_place_of_death %>%
  group_by(place) %>%
  filter(max(fract) > 1) %>%
  ggplot(aes(place, fract, fill = cause)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_nejm() +
  labs(title = paste0("Place of death [data up to ", data_up_to, "]"),
       y = "Fraction of reported deaths",
       fill = "Cause of death:") +
  theme(axis.title.x = element_blank(), text = element_text(size = 16))
```

```{r, fig.width=14, fig.height=7, eval=FALSE, echo=FALSE}
place_of_death_long %>% 
  filter(country == "England and Wales") %>%
  mutate(place = gsub("Hospital \\(acute or community, not psychiatric\\)", 
                      "Hospital", place),
         place = gsub("Other communal establishment", 
                      "Other communal\nestablishment", place),
         week_ = ymd("2020-01-03") + 7 * (week-1),
         number_of_deaths = as.numeric(number_of_deaths)) %>%
  filter(place %in% c("Care Home", "Home", "Hospice", "Hospital")) %>%
  ggplot(aes(week_, number_of_deaths, fill = place)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_lancet() +
  facet_grid(cause~place) +
  labs(title = paste0("Place of death [data up to ", data_up_to, "]"),
       y = "Number of reported deaths",
       fill = "Place of death:") +
  theme(axis.title.x = element_blank(),
        text = element_text(size = 16)) +
  scale_x_date(breaks = ymd("2020/03/13", "2020/04/03"), date_labels = "%b")
```

We see increase in almost all places of death. However, COVID-19 accounts only for some of the increase. 

_My speculation:_ Is this because there was no test made? I find it unlikely, as the deaths reported to ONS don't have to be verified by testing to be classified as COVID-19 related. It might be, for example, because [people are scared to go to the hospitals.](https://www.washingtonpost.com/health/patients-with-heart-attacks-strokes-and-even-appendicitis-vanish-from-hospitals/2020/04/19/9ca3ef24-7eb4-11ea-9040-68981f488eed_story.html) 

Also, as have been already reported, the outbreak in care homes started later, hence we are seeing an increase in COVID-19 related deaths in those facilities. 

Below we see that number of deaths with no mention of COVID-19 decreased in the hospitals. I hypothesize it potentially can be explained by increase in deaths in other places, less road accidents and less scheduled medical interventions.

**Note:** _Many thanks to [German](https://mobile.twitter.com/not_a_reptiloid) (and [David Spiegelhalter plots](https://mobile.twitter.com/d_spiegel/status/1255148797333655552/photo/1)) for making me realise a mistake in the earlier iteration of the following plot!_

```{r, fig.width=10, fig.height=5}
place_of_death_long %>% 
  filter(country == "England and Wales") %>%
  mutate(place = gsub("Hospital \\(acute or community, not psychiatric\\)", 
                      "Hospital", place),
         place = gsub("Other communal establishment", 
                      "Other communal\nestablishment", place),
         week_ = ymd("2020-01-03") + 7 * (week-1),
         number_of_deaths = as.numeric(number_of_deaths)) %>%
  pivot_wider(names_from = "cause", values_from = number_of_deaths) %>% 
  mutate(`No COVID-19 mentioned` = `Total deaths` - `COVID-19 deaths`) %>%
  select(-`Total deaths`) %>%
  pivot_longer(names_to = "cause", values_to = "number_of_deaths", 
               -c(place, country, week, week_)) %>%
  group_by(place) %>%
  mutate(max_deaths = max(number_of_deaths)) %>%
  ungroup() %>%
  mutate(place = reorder(place, -max_deaths)) %>%
  filter(place %in% c("Care Home", "Home", "Hospital")) %>%
  ggplot(aes(week_, number_of_deaths, fill = cause)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = pal_nejm()(6)[c(1,6)]) +
  facet_wrap(~place) +
  labs(title = paste0("Place of death [data up to ", data_up_to, "]"),
       y = "Number of reported deaths",
       fill = "Place of death:") +
  theme(axis.title.x = element_blank(),
        text = element_text(size = 16)) +
  scale_x_date(breaks = ymd("2020/03/13", "2020/04/03"), date_labels = "%b")
```


### Discrepancies between the reported numbers

The numbers reported by the government are lower than the actual number of deaths each day. Some of that comes from the delay in confirming the cause of death. It might be the case that the `Place of death` plot will change as the data will be updated. 

All the comparisons between previous years (in the earlier parts of this notebook) are done based on figures from **ONS deaths by date of registration – registered by `r format(data_up_to, '%d %b')`** set for compatibility with earlier datasets. 

```{r, fig.width=13, fig.height=7}
col_scheme <- c("#E18727", "#849460", "#56603F", "#20854E")
names(col_scheme) <- unique(actual_vs_reported$source)[1:4]

label_postions <- 
  actual_vs_reported %>% 
  filter(source %in% names(col_scheme)) %>%
  group_by(source) %>%
  mutate(max = max(number_of_deaths)) %>%
  arrange(Date) %>%
  mutate(deaths_on_that_day = number_of_deaths - lag(number_of_deaths),
         max_on_day = max(deaths_on_that_day, na.rm = TRUE)) %>%
  ungroup() %>%
  filter(Date == lubridate::ymd("2020/03/6")) %>%
  arrange(-max) %>%
  mutate(fixed_position_cum = max(max) - 0.07 * max(max) * 1:length(col_scheme),
         fixed_position_daily = max(max_on_day) - 0.07 * max(max_on_day) * 1:length(col_scheme))

discrepancies <-
  actual_vs_reported %>%
  filter(source %in% names(col_scheme)) %>%
  ggplot(aes(x = Date, y = number_of_deaths, color = source, shape = source)) +
  geom_smooth(se = FALSE) +
  geom_point(color = "black") +
  geom_text(data = label_postions, size = 7, hjust = 0,
            aes(x = Date, y = fixed_position_cum, label = source)) + 
  geom_point(data = label_postions, color = "black",
             aes(x = as_datetime(ymd(Date)-0.5), 
                 y = fixed_position_cum, shape = source)) +
  theme(legend.position = "right") +
  #scale_color_nejm() +
  scale_color_manual(values = col_scheme) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))+
  theme(axis.title.x = element_blank(), text = element_text(size = 16),
        legend.position = "none") +
  labs(color = "Source", shape = "Source", 
       y = "Cummulative number of deaths",
       title = "Delays in the reported numbers of COVID-19 deaths in England and Wales")

discrepancies
# (discrepancies + theme(legend.position = "none") + labs(title = "[linear scale]")) / 
#   (discrepancies + scale_y_log10() + labs(title = "[log scale]")) + 
#   plot_annotation(title =  "Discrepancies between the reported numbers")
```

Looking at the daily death rates it becomes clear that there is a significant lag in the data. And the drops in the registered deaths on the weekends (March 21 & 22, 28 & 29, April 4 & 5, Easter weekend, and so on). 


```{r, fig.width=13, fig.height=7}
actual_vs_reported %>%
  filter(source %in% names(col_scheme)) %>%
  group_by(source) %>%
  arrange(Date) %>%
  mutate(deaths_on_that_day = number_of_deaths - lag(number_of_deaths)) %>%
  ggplot(aes(x = Date, y = deaths_on_that_day, color = source, shape = source)) +
  geom_smooth(se = FALSE) +
  geom_point(color = "black") +
  geom_text(data = label_postions, size = 7, hjust = 0,
            aes(x = Date, y = fixed_position_daily, label = source)) +
  geom_point(data = label_postions, color = "black",
            aes(x = as_datetime(ymd(Date)-0.5), 
                y = fixed_position_daily, shape = source)) +
  scale_color_manual(values = col_scheme) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))+
  labs(color = "Source", shape = "Source", 
       y = "Number of deaths on the day",
       title = "Discrepancies between the reported numbers of COVID-19 deaths in England and Wales") +
  theme(axis.title.x = element_blank(), legend.position = "none",
        text = element_text(size = 16))
```

### Age distribution

Looking at the deaths, Males are more affected than Females. How this plot looks like for cases? I.e. are men more susceptible? 

I also wanted to compare the COVID-19 age and sex distribution with the UK population structure. I ended up comparing it to even more adequate data - to the deaths occuring in UK this year, before COVID-19. This will allows to put the age structure of COVID-19 into context.

```{r, eval=FALSE, echo=FALSE}
max_axis <- 
  age_distribution %>%
  group_by(sex, age_group) %>%
  summarise(number_of_deaths = sum(number_of_deaths)) %>%
  ungroup() %>%
  filter(sex != "Both") %>%
  pull(number_of_deaths) %>%
  max()
max_axis <- round(max_axis / 100, 0) * 100 + 25
break_axis <- sort(c(-round(seq(0, max_axis, length.out = 5)), 
                     round(seq(0, max_axis, length.out = 5))))
break_axis <- round(break_axis / 50) * 50

covid_age <- 
  age_distribution %>%
  mutate(age_group = factor(age_group, levels = rev(age_levels))) %>%
  filter(sex != "Both") %>%
  mutate(nod = ifelse(sex == "Females", number_of_deaths, -number_of_deaths)) %>%
  group_by(age_group) %>%
#  filter(sum(number_of_deaths) > 0) %>%
  ggplot(aes(x = age_group, y = nod, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_npg() +
  scale_y_continuous(breaks = break_axis,
                     labels = abs(break_axis)) +
  labs(x = "Age groups",
       y = paste0("Number of deaths [data up to ", data_up_to, "]"), 
       fill = "Sex") +
  coord_flip() +
  theme(legend.position = "right")

```


```{r, fig.width=10, fig.height=7}
before_covid <- ymd("2020-02-28")

age_distribution_bound <-
  bind_rows(age_distribution_all %>%
            mutate(type = "Before COVID-19"),
          age_distribution %>%
            mutate(type = "COVID-19")) %>%
  filter(sex != "Both") %>%
  group_by(sex, age_group, type) %>%
  summarise(number_of_deaths = sum(number_of_deaths)) %>%
  group_by(type) %>%
  mutate(all_deaths = sum(number_of_deaths)) %>%
  group_by(sex, type, age_group) %>%
  mutate(fraction_of_deaths = number_of_deaths/all_deaths * 100) %>%
  ungroup()

age_distribution_bound %>%
  mutate(age_group = factor(age_group, levels = rev(age_levels)),
         nod = ifelse(sex == "Females", number_of_deaths, -number_of_deaths)) %>%
  group_by(age_group) %>%
  filter(sum(number_of_deaths) > 0) %>%
  ggplot(aes(x = age_group, y = nod, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_npg() +
  facet_wrap(~type, scales = "free_x") +
  # scale_y_continuous(breaks = break_axis,
  #                    labels = abs(break_axis)) +
  labs(x = "Age groups",
       y = paste0("Number of deaths\n[Data up to:",
                  " before COVID-19 -- ", before_covid, 
                  "; COVID-19 -- ", data_up_to, 
                  "]"), 
       fill = "Sex",
       title = ) +
  coord_flip() +
  theme(legend.position = "bottom",
        text = element_text(size = 16))
```

This is even more visible if we would compare the fraction of all deaths in each age group and with respect to sex.

```{r, fig.width=10, fig.height=7}
age_distribution_bound %>% 
  group_by(age_group) %>% 
  filter(sum(fraction_of_deaths) > 5) %>% 
  ungroup() %>%
  ggplot(aes(age_group, fraction_of_deaths, fill = type)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  facet_wrap(~sex) +
  scale_fill_jama() +
  labs(title = "Fraction of all deaths with respect to age group and sex",
       subtitle = "age groups contirbuting to fewer than 5% filtered out") + 
  theme(axis.title = element_blank(), legend.title = element_blank(),
        text = element_text(size = 16))
```


