---
title: "COVID-19 mortality in England and Wales"
subtitle: "ONS data up to 27/03/2020"
author: Kasia Kedzierska
output:
  html_document:
    df_print: paged
    code_folding: hide
    theme: paper
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 3
---
## Motivation

As probably many of the data scientist I wanted to ease my anxiety caused by the COVID-19, and  for some reason analysing COVID-19 data is doing that job. I am a **computational biologists** and all the plots I am making here are just *observations*. I am not claiming any conclusions, nor do I do any predictions. All the code is open source, and free for anyone to use, available on [github](https://github.com/kzkedzierska/COVID-19_England_Wales).

Also, very warm thanks to [Kaspar](https://kasparmartens.rbind.io/) for his help with decisions regarding color schemes and his pointers.

## Setup 



Loading up the necessary packages (installing them if they are missing). Setting up defaults and tweaking the plotting defaults.

```{r, warning=FALSE, error=FALSE, message=FALSE}
# Load packages
cran_pkgs <- c("tidyverse", "ggsci", "patchwork", "readxl", "knitr", "ggforce")
source("./scripts/load_pkgs.R")
load_packages(cran_pkgs = cran_pkgs)

# Set defaults for plotting
theme_set(theme_bw() + theme(legend.position = "bottom"))

# Don't print warnings
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```
## Data 

Data is coming from [Office for National Statistics, *Deaths registered weekly in England and Wales, provisional* dataset](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/deaths/datasets/weeklyprovisionalfiguresondeathsregisteredinenglandandwales). For now I am downloading the data manually into the `data` directory. It would be great to download it directly.

### Weekly figures 

Both 2020 and 2019 figures for comparison of the trend. The 2020 will be changed each week, while 2019 remains stable. 

*Note:* Unfortunately due to formatting of the data the following code is not entirely reproducible as there are parts of it that would have to be adjusted with each newiteration of the spreadsheet. 

```{r, warning=FALSE, message=FALSE}
df_2019 <- 
  read_xls("./data/publishedweek522019.xls", 
           sheet = "Weekly figures 2019", skip = 3)[c(4,5,9),-c(1,2)] %>%
  mutate(value = c("total_deaths", "total_deaths_ave5", "respiratory_deaths")) %>%
  mutate_at(vars(-value), as.numeric) %>% 
  pivot_longer(names_to = "week", 
               values_to = "number_of_deaths", -value) %>%
  mutate(year = "2019")

current_2020_spreadsheet <- "./data/publishedweek1320201.xlsx"
sheets_2020 <- readxl::excel_sheets(current_2020_spreadsheet)
df_2020_ <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[5], skip = 4)[c(4,6,9, 10),-c(1,2)] %>%
  mutate(value2 = c("total_deaths", "total_deaths_ave5",
                   "respiratory_deaths", "covid_deaths"),
         value = c("total_deaths", "total_deaths_ave5",
                   "respiratory_deaths", "respiratory_deaths")) %>%
  mutate_at(vars(-value, -value2), as.numeric) %>% 
  pivot_longer(names_to = "week", 
               values_to = "number_of_deaths", 
               -c(value, value2))

df_2020 <- 
  df_2020_ %>% 
  select(-value2) %>%
  mutate(year = "2020") %>%
  filter(!is.na(number_of_deaths))

df_all <- bind_rows(list(df_2019, df_2020)) %>%
  mutate(week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  group_by(value, week, week_, year) %>%
  summarise(number_of_deaths = sum(number_of_deaths)) %>%
  ungroup()
```

### COVID-19 specific datasets

*Note:* Total number of deaths in `Covid-19 - Place of occurence` sheet is off by a factor of 10 in the data published on 7th of April.

```{r}
place_of_death <- 
  read_xlsx(current_2020_spreadsheet, 
            sheet = sheets_2020[9], skip = 4)[1:7,1:3] %>%
  rename(place = `...1`, 
         `All` = `Number of deaths`, 
         `COVID-19` = `Number of COVID-19 deaths`) %>%
  filter(place != "Total") %>%
  pivot_longer(names_to = "cause", 
               values_to = "number_of_deaths",
               -place)

actual_vs_reported <-
  read_xlsx(current_2020_spreadsheet, sheet = sheets_2020[7], skip = 4, 
            col_types = c("date", "numeric", 
                          "numeric", "numeric", "numeric"))[1:23, 1:5] %>%
  pivot_longer(names_to = "source",
               values_to = "number_of_deaths",
               -Date)

age_distribution <- 
  read_xlsx(current_2020_spreadsheet, sheet = sheets_2020[6], 
            skip = 4)[c(7:13,16:22,25:31), -1] %>%
  mutate(sex = c(rep("Both", 7), rep("Males", 7), rep("Females", 7))) %>%
  rename(age_group = `...2`) %>%
  mutate_at(vars(-age_group, -sex), as.numeric) %>%
  pivot_longer(names_to = "week",
               values_to = "number_of_deaths",
               cols = -c(age_group, sex)) %>%
  group_by(sex, week) %>%
  filter(sum(number_of_deaths) > 0)

age_levels <- age_distribution$age_group %>% unique()
age_distribution <-
  age_distribution %>%
  ungroup() %>%
  mutate(week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  mutate(age_group = factor(age_group, levels = age_levels))
```


## Mortality in the context of past years

### All weekly deaths in England and Wales

Still very early, but looks like the overall weekly numbers are higher than last year, and the past 5 years.

```{r, fig.height=8, fig.width=7}
total <- 
  df_all %>%
  filter(value == "total_deaths") %>%
  ggplot(aes(x = week_, y = number_of_deaths, fill = year)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_smooth(data = filter(df_all, value == "total_deaths_ave5" & year == "2019"),
              color = "grey60", se = FALSE) +
  facet_zoom(xy = week_ < lubridate::ymd("2020-03-28"), 
             horizontal = FALSE, show.area = FALSE) +
  labs(y = "Number of deaths", 
       title = "Weekly total deaths [grey = 5 year average]") + 
  theme(axis.title.x = element_blank()) +
  scale_fill_aaas()
total
```

### Respiratory deaths

COVID-19 deaths are increasing the number of repiratory deaths. 

```{r, fig.height=8, fig.width=7}
respiratory <-
  df_all %>%
  filter(value == "respiratory_deaths") %>%
  ggplot(aes(x = week_, y = number_of_deaths, 
             group = year, color = year, shape = year)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_zoom(xy = week_ < lubridate::ymd("2020-03-28"), 
             horizontal = FALSE, show.area = FALSE) +
  scale_color_aaas() +
  labs(y = "Number of deaths", title = "Weekly respiratory deaths") + 
  theme(axis.title.x = element_blank()) +
  coord_cartesian(ylim = c(0, 2510))
respiratory
```

However, the difference is not yet visible on the cummulative scale. 

```{r, fig.height=8, fig.width=7}
cumulative_respiratory <- 
  df_all %>%
  arrange(week) %>%
  group_by(value, year) %>%
  mutate(cum = cumsum(number_of_deaths)) %>%
  filter(value == "respiratory_deaths") %>%
  ungroup() %>% 
  ggplot(aes(x = week_, y = cum, color = year, shape = year)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  facet_zoom(xy = week_ < lubridate::ymd("2020-03-28"), 
             horizontal = FALSE, show.area = FALSE) +
  scale_color_aaas() +
  labs(y = "Number of deaths", shape = "Year",
       color = "Year", title = "Cumulative respiratory deaths") + 
  theme(axis.title.x = element_blank()) 
cumulative_respiratory
```

COVID-19 deaths increase exponentailly, while the other respiratory deths seem to be stable.

```{r}
df_2020_ %>%
  filter(value == "respiratory_deaths") %>%
  mutate(value2 = gsub("respiratory_deaths", "other_respiratory_deaths", value2),
         week = as.numeric(week), 
         week_ = lubridate::ymd("2020-01-03") + 7 * (week-1)) %>%
  filter(!is.na(number_of_deaths)) %>%
  ggplot(aes(x = week_, y = number_of_deaths, color = value2, shape = value2)) +
  geom_point() +
  geom_smooth(se = FALSE) +
  scale_color_uchicago() +
  labs(y = "Number of deaths", title = "Weekly respiratory deaths",
       color = "Cause of death", shape = "Cause of death") + 
  theme(axis.title.x = element_blank(), legend.position = "bottom")
```

```{r, fig.width=7, fig.height=12, eval=FALSE}
total / respiratory / cumulative_respiratory + plot_layout(guides = "collect")
ggsave("./plots/up_to200327.png")
```

## COVID-19 specific stats

### Place of deaths

As of 27 of March, there is not a lot of deaths reported outside of the hospitals. This however can change, as we get more data. 

```{r}
place_of_death %>% 
  mutate(place = gsub("Hospitals \\(acute or community not psychiatric\\)", 
                      "Hospitals", place),
         place = gsub("Other communal establishments", 
                      "Other communal\nestablishments", place)) %>%
  group_by(cause) %>%
  mutate(fract = round(number_of_deaths / sum(number_of_deaths) * 100, 2)) %>%
  group_by(place) %>%
  filter(max(fract) > 1) %>%
  ggplot(aes(place, fract, fill = cause)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_nejm() +
  labs(title = "Place of death",
       y = "Fraction of reported deaths",
       fill = "Cause of death:") +
  theme(axis.title.x = element_blank(), text = element_text(size = 16))
```

### Discrepancies between  reported numbers

The numbers reported by the government are lower than the actual number of deaths each day. Some of that comes from the delay in confirming the cause of death. It might be the case that the `Place of death` plot will change as the data will be updated. 

All the comparisons between previous years (in the earlier parts of this notebook) are done based on figures from `ONS deaths by date of registration – registered by 27th March` set for compatibility with earlier datasets.

```{r, fig.width=13, fig.height=7}
col_scheme <- c("#E18727", "#849460", "#56603F", "#20854E")
                #"#20854E", "#008B6C", "#009093")
names(col_scheme) <- unique(actual_vs_reported$source)

label_postions <- 
  actual_vs_reported %>% 
  filter(Date == lubridate::ymd("2020/03/5")) %>%
  arrange(number_of_deaths) %>%
  mutate(fixed_position_cum = 1000 + 100 * 1:4,
         fixed_position_daily = 150 + 15 * 1:4)

discrepancies <-
  actual_vs_reported %>%
  ggplot(aes(x = Date, y = number_of_deaths, color = source, shape = source)) +
  geom_smooth(se = FALSE) +
  geom_point(color = "black") +
  geom_text(data = label_postions, size = 7, hjust = 0,
            aes(x = Date, y = fixed_position_cum, label = source)) +
  #scale_color_nejm() +
  scale_color_manual(values = col_scheme) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))+
  labs(color = "Source", shape = "Source", 
       y = "Cumulative number of deaths") +
  theme(axis.title.x = element_blank(), text = element_text(size = 16),
        legend.position = "none") 

discrepancies + 
  labs(title = "Discrepancies between the reported numbers of COVID-19 deaths in England and Wales")
# (discrepancies + theme(legend.position = "none") + labs(title = "[linear scale]")) / 
#   (discrepancies + scale_y_log10() + labs(title = "[log scale]")) + 
#   plot_annotation(title =  "Discrepancies between the reported numbers")
```

Looking at the daily death rates it becomes clear that there is a significant lag in the data.

```{r, fig.width=13, fig.height=7}
actual_vs_reported %>%
  group_by(source) %>%
  arrange(Date) %>%
  mutate(deaths_on_that_day = number_of_deaths - lag(number_of_deaths)) %>%
  ggplot(aes(x = Date, y = deaths_on_that_day, color = source, shape = source)) +
  geom_smooth(se = FALSE) +
  geom_point(color = "black") +
  geom_text(data = label_postions, size = 7, hjust = 0,
            aes(x = Date, y = fixed_position_daily, label = source)) +
  scale_color_manual(values = col_scheme) +
  guides(color = guide_legend(nrow = 2, byrow = TRUE))+
  labs(color = "Source", shape = "Source", 
       y = "Number of deaths on the day") +
  theme(axis.title.x = element_blank(), legend.position = "none",
        text = element_text(size = 16)) 
  
```



### Age distribution

There is more deaths in among Males than Females. I

```{r}
max_axis <- 
  age_distribution %>%
  filter(sex != "Both") %>%
  pull(number_of_deaths) %>%
  max()
max_axis <- round(max_axis / 100, 0) * 100 + 25

age_distribution %>%
  mutate(age_group = factor(age_group, levels = rev(age_levels))) %>%
  filter(week == max(week), sex != "Both") %>%
  mutate(nod = ifelse(sex == "Females", number_of_deaths, -number_of_deaths)) %>%
  group_by(age_group) %>%
  filter(sum(number_of_deaths) > 0) %>%
  ggplot(aes(x = age_group, y = nod, fill = sex)) +
  geom_bar(stat = "identity") +
  scale_fill_npg() +
  scale_y_continuous(breaks = seq(-max_axis, max_axis, 25),
                     labels = abs(seq(-max_axis, max_axis, 25))) +
  labs(x = "Age groups",
       y = "Cumulative number of deaths", fill = "Sex") +
  coord_flip() +
  theme(legend.position = "right")

```






